Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models

### 2022-11-27 manuscript/manuscript.rmd
adding readme page for spatiotemporal animation

---
title: Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
author: 
  - name: "Christian Testa"
    institute: "1"
    email: "ctesta@hsph.harvard.edu"
    correspondence: true
institute: 
  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
abstract: |
  We present three types of applications of generalized additive models (GAMs) to
  COVID-19 mortality rates in the US for the purpose of advancing methods to
  document inequities with respect to which communities suffered disproportionate
  COVID-19 mortality rates. First, GAMs can be used to describe the changing
  relationship between COVID-19 mortality and county-level covariates
  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
  be used to perform spatiotemporal smoothing that pools information over time and
  space so that county level mortality rates that tend to be noisy due to small
  population counts or stochasticity can be summarized by a smooth, dynamic latent
  surface describing the mortality risk associated with geographic locations over
  time. Third, estimation of the associations between county-level covariates and
  COVID-19 mortality after controlling for spatiotemporal effects allow for the
  distinguishing of what patterns in COVID-19 mortality were more plausibly due to
  geography than variation in county-level covariates. Each of these approaches
  provides a valuable approach and perspective to documenting inequities in
  COVID-19 mortality by addressing the question of which populations have suffered
  the worst burden of COVID-19 mortality taking into account the nonlinear
  spatial, temporal, and social patterning of disease.
  \newline

@@

As we enter the third winter with the novel coronavirus disease 
COVID-19 in the United States, evidence documenting the intense disparities in
COVID-19 mortality rates comparing socially advantaged and disadvantaged
populations continues to mount. Eliminating inequities in health outcomes has
been stated as a major policy goal of the Biden administration
[@the_white_house_executive_2021; @the_white_house_advancing_2022], representing
a revitalized focus on health justice and underlining the importance of adequate
data reporting systems that measure the prevalent health inequities. To this
end, we use generalized additive models (GAMs) as a flexible regression
framework to illustrate the evolving roles and relationships sociodemographic,
geographic, and economic conditions have with respect to trends in COVID-19
mortality. The code, data, and documentation necessary to reproduce the analyses
contained in this paper are online and free to access at <https://github.com/ctesta01/covid.gradient.estimation>.

@@

Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
The figure summarizes each division's median mortality rates
weighted by county population size. Quantile ranges are included to illustrate
the range of mortality rates observed, again weighted by county population size.
Weighting based on county population size is used as opposed to equal weighting
for calculating the median and quantile ranges so that in divisions where low
population counties predominate, mortality rate estimates that are noisy due to
small population counts are not weighted equally to more reliable mortality rate
estimates from larger population counties. Notably, the mortality associated
with the early surge of cases starting in NYC and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

@@

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61%-100%) compared
to counties in the lowest quintile of percent People of Color (0%-17.2%)
[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
barriers in COVID-19 prevention and care: 1) originally requiring New York
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve [@okonkwo_covid-19_2021]. It's not just a hypothetical 
that those with more structurally enfranchised privileges have been more able to
mitigate their risk of negative health outcomes associated with COVID-19 
— during February 1 to April 1, 2020, New York City residents who lived in more affluent 
neighborhoods were more likely to have left the city, while New York City residents
from more Black and Hispanic neighborhoods were more likely to continue working [@coven_disparities_2020].

@@

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at "}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

## Evaluating Spatial and Temporal Autocorrelation

@@

```{r figure 5, fig.cap = ""}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

>>>>

---
title: Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
author: 
  - name: "Christian Testa"
    institute: "1"
    email: "ctesta@hsph.harvard.edu"
    correspondence: true
institute: 
  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
abstract: |
  I present three types of applications of generalized additive models (GAMs) to
  COVID-19 mortality rates in the US for the purpose of advancing methods to
  document inequities with respect to which communities suffered disproportionate
  COVID-19 mortality rates at specific times during the first three years of the 
  pandemic. First, GAMs can be used to describe the changing
  relationship between COVID-19 mortality and county-level covariates
  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
  be used to perform spatiotemporal smoothing that pools information over time and
  space to address statistical instability due to small
  population counts or stochasticity can be summarized by a smooth, dynamic latent
  surface describing the mortality risk associated with geographic locations over
  time. Third, estimation of COVID-19 mortality associations with county-level covariates
  conditional on smooth spatiotemporal variation allows for more rigorous consideration
  of how socio-environmental contexts and policies may have impacted COVID-19 mortality.
  Each of these approaches
  provides a valuable perspective to documenting inequities in
  COVID-19 mortality by addressing the question of which populations have suffered
  the worst burden of COVID-19 mortality taking into account the nonlinear
  spatial, temporal, and social patterning of disease.
  \newline

@@

As we enter the third winter with the novel coronavirus disease 
COVID-19 in the United States, evidence documenting the intense disparities in
COVID-19 mortality rates comparing socially advantaged and disadvantaged
populations continues to mount. Eliminating inequities in health outcomes has
been stated as a major policy goal of the Biden administration
[@the_white_house_executive_2021; @the_white_house_advancing_2022], representing
a revitalized commitment to health equity and underlining the importance of adequate
data reporting systems that report timely estimates of prevalent health inequities. To this
end, I use generalized additive models (GAMs) as a flexible regression
framework to illustrate the evolving roles and relationships sociodemographic,
geographic, and economic conditions have with respect to trends in COVID-19
mortality. The code, data, and documentation necessary to reproduce the analyses
contained in this paper are online and free to access at <https://github.com/ctesta01/covid.gradient.estimation>.

@@

Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
The figure summarizes each division's median mortality rates
weighted by county population size. Quantile ranges are included to illustrate
the range of mortality rates observed, again weighted by county population size.
Notably, the mortality associated
with the early surge of cases starting in NYC and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

@@

Even since the beginning of the COVID-19 outbreak in the US, data reflected
sharp inequities in mortality rates.  During January 22nd to May 5th 2020,
county COVID-19 mortality rates were 4.94 (95% CI 4.78, 5.09) times higher in
counties in the highest quintile of percent People of Color (61%-100%) compared
to counties in the lowest quintile of percent People of Color (0%-17.2%)
[@chen_revealing_2021].  This was not wholly unanticipated: as COVID-19 was
beginning to take off in the US, some were already calling attention to the fact
that COVID-19 threatened to exacerbate existing disparities
[@kim_covid-19_2020]. Kim, Marrast, and Conigliaro noted three structural
barriers in COVID-19 prevention and care: 1) originally requiring New York
residents to have a doctor's prescription for a COVID-19 test reduced the
opportunity for healthcare for People of Color as they are less likely to have a
primary care provider; 2) drive-through testing made testing disadvantaged those
who rely on public transportation; 3) quarantining at home while waiting the
7-10 days originally required for COVID-19 test results to come back posed an
economic and social challenge that many in already financially difficult
situations may not have been able to take on [-@kim_covid-19_2020]. Others noted
yet more reasons why COVID-19 threatened to worsen an already inequitable
healthcare landscape in the US: those who reside in prisons and jails,
immigrants and undocumented people, people with disabilities, and people
experiencing homeless all face additional challenges in seeking and getting the
healthcare they deserve [@okonkwo_covid-19_2021]. Despite stay-at-home orders
that were prevalent in many states [@moreland_timing_2020], workers functioning in 
capacities essential to the functioning of critical infrastructure operations (later
termed "essential workers") were exposed to heightened risk of COVID-19 transmission 
[@national_conference_of_state_legislatures_covid-19_2021; @wei_risk_2022; @national_bureau_of_economic_research_measuring_2021; @the_lancet_plight_2020; @hanage_covid-19_2020].
It's not just a hypothetical that those with more structurally enfranchised
privileges have been more able to mitigate their risk of negative health
outcomes associated with COVID-19 — during February 1 to April 1, 2020, New York
City residents who lived in more affluent neighborhoods were more likely to have
left the city, while New York City residents from more Black and Hispanic
neighborhoods were more likely to continue working [@coven_disparities_2020].

@@

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at [https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md](https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md)"}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

## Application 2: Spatiotemporal Smoothing

By fitting GAMs with a tensor-product smoothing term on latitude, longitude, and
time we can estimate a spatiotemporally smoothed trend in COVID-19 mortality. To
do this, the GAM is constructed similarly as in Application 1 but with the
formula specified as `deaths ~ te(latitude, longitude, time, d=c(2,1))` where
`d=c(2,1)` indicates that `latitude` and `longitude` share the same dimensions
(i.e., both are spatial and in units of degrees) while the `time` data are in separate
units.

Using GAMs to present spatiotemporally smoothed estimates of COVID-19 mortality
allows for the synthesis of trends over time. Whereas the raw rates of COVID-19
mortality are noisy due to 72% of the land area in the US being classified as
rural and low population density, the spatiotemporally smoothed GAM estimates
use population weighting via the offset specified to estimate a latent surface
that represents localized mortality rate averages in spatial coordinates and in
time.

```{r crude-vs-smoothed, fig.cap = "Comparison of crude vs. smoothed mortality rates. A) Crude mortality rates per 100,000 person-years in January 2022. B) Smoothed mortality rates from a GAM applied to mortality rates from January 2022."}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/side_by_side_crude_and_smoothed.png"))
```

```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at []()"}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

## Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation

### Evaluating Spatial and Temporal Autocorrelation

@@
